---
title: "CH11: Inference with Mathematical Models"
subtitle: "Working With the Normal Distribution"
output:
  html_document:
    theme: yeti
    highlight: textmate
    toc: true
    toc_depth: 4
---

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE,
  comment = NA 
)
```

```{r}
library(tidyverse)
library(openintro)
```

The `normTail` function: https://openintrostat.github.io/openintro/reference/normTail.html

The normal distribution functions in `R`: https://rdrr.io/r/stats/Normal.html

## SAT and ACT Scores Examples

SAT scores follow a nearly normal distribution with a mean of 1500 points and a standard deviation of 300 points. ACT scores also follow a nearly normal distribution with mean of 21 points and a standard deviation of 5 points. Suppose Mel scored 1800 points on their SAT and Sian scored 24 points on their ACT. Who performed better?

Note: This refers to the three-section version of the SAT (used in 2005–2015), where Math, Critical Reading, and Writing were each worth 800 points.

### Mel's SAT Percentile, Z-score, and Plots
```{r}
melsat  <- 1800
satmean <- 1500
satsd   <- 300
```

```{r}
# create the raw score plot

```

```{r}
# find the raw score percentile

```

```{r}
# calculate the z-score

```

```{r}
# create the standardized score plot

```

```{r}
# find the z-score from the percentile

```

### Sian's ACT Percentile, Z-score, and Plots
```{r}
sianact <- 24
actmean <- 21
actsd   <- 5
```

```{r}
# create the raw score plot

```

```{r}
# find the raw score percentile

```

```{r}
# calculate the z-score

```

```{r}
# create the standardized score plot

```

```{r}
# find the z-score from the percentile

```

### Mel's ACT Percentile

Suppose Mel also took the ACT and scored at the exact same percentile as his SAT score. What would be his ACT raw score?
```{r}
# plot of the percentile

```

```{r}
# raw score calculation

```

### Shannon's SAT Performance

Shannon is a randomly chosen SAT taker; nothing is known about their aptitude. What is the probability they score at least 1630?
```{r}
# plot of the raw scores

```

```{r}
# raw probability calculation

```

```{r}
# z-scores calculations

```

```{r}
# plot of the z-scores

```

```{r}
# z-scores probability calculation

```

### Random Score Generation

Generate scores for 100 randomly selected SAT-takers.
```{r}

```

