---
title: "EDA &#9654; Categorical Data<br>Introducing Factors"
author: "Author: Jill E. Thomley"
date: '`r format(Sys.time(), "%A, %B %d, %Y @ %I:%M %p")`'
output: 
  ioslides_presentation:
    logo: images/logoASU.jpg
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA)
library(tidyverse)
```


## Before We Begin...

These slides are not meant to be standalone information. You should take notes to flesh out the contents. I recommend that you create an R Markdown document where you can combine information and code from the slides and your own additional notes and explorations to make connections.

**Related Materials**

* CH14 of [R for Data Science](https://r4ds.had.co.nz/strings.html)
* CH15 of [R for Data Science](https://r4ds.had.co.nz/factors.html)
* [Understanding Factors](http://monashbioinformaticsplatform.github.io/2015-09-28-rbioinformatics-intro-r/01-supp-factors.html) from the Software Carpentry R course.


## What is a Factor in `R`?

By default, `R` displays the possible values of a character vector in ascending alphabetical order. When we summarize, visualize, or model with categorical variables, we often want their categories to appear in a different order. 

Structurally, factor variables are stored as unique integers with attached labels that are used for display. This is similar to how logical vectors display `TRUE` and `FALSE` but are stored as 0's and 1's "under the hood". 

The set of possible values of a factor are are known as *levels*.

The `factor()` function is used to create and modify factors in base `R` and the tidyverse contains a package called forcats that offers other useful functions.


## Revisiting An Example

The [STT3850majors.csv](https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/stt3850majors.csv) dataset (click the link for the URL to the CSV file) contains information about students in Dr. Thomley's past STT3850 classes. 

```{r echo=FALSE}
students <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/Instructor/stt3850majors.csv")
```

* College = ASU college the student is in (e.g., Arts & sciences)
* Year = student's class year (e.g., freshman, sophomore)
* BannerMajor = student's major as listed in banner
* CodedMajor = Dr. Thomley's recoded list of majors
* Semester = the semester when the student took STT3850
* AY = academic year (e.g., 2018-2019)


## Dataset Structure

What are the current vector types of variables in the dataset?

```{r}
glimpse(students)
```

All the variables are stored as character (`<chr>`) vectors.


## Class Years

Notice the categories for `Year` are not in the standard order. We should probably convert the character variable to a factor. Then we can choose an order other than `R`'s alphabetical default.

```{r}
students %>% 
  count(Year)
```


## Converting to a Factor

We know from our previous table that there are five categories or levels. We will create a vector that puts all of the levels in the order we want them. Then create our factor using `factor()` to replace the original version.

```{r}
students <- students %>%
  mutate(Year = factor(Year,
    levels = c(
      "Freshman",
      "Sophomore",
      "Junior",
      "Senior",
      "Graduate Certificate Program"
    )
  ))
```


## Dataset Structure

Since we saved the factor using the original dataset and variable names, the original character variable has been replaced.

```{r}
glimpse(students)
```

Now year is stored as a factor (`<fct>`) vector.


## 

```{r}
# tidyverse lets us mutate() multiple variables in one statement
studentsA <- students %>%
  mutate(
    Year = factor(Year,
      levels = c(
        "Freshman",
        "Sophomore",
        "Junior",
        "Senior",
        "Graduate Certificate Program"
      )
    ),
    CodedMajor = factor(CodedMajor,
      levels = c(
        "Computer Science",
        "Mathematical Sciences",
        "Other Sciences",
        "Other"
      )
    )
  )
```


## Now the Years Table Again

The class years are now in the customary order, as determined by the imposed factor structure. 

```{r}
studentsA %>% 
  count(Year)
```


## And the Majors Table

Now the catch-all `Other` category is last, as is typically seen in summaries, because of the imposed structure. There are also *ordered factors* that have additional structure for higher level analyses. We do not need those for EDA.

```{r}
studentsA %>% 
  count(CodedMajor) %>% 
  mutate(percents = 100 * prop.table(n))
```


## Renaming Using the `labels =` Option

You can use the `labels =` parameter in `factor()` to create alternate names that will be displayed for the various factor levels in plots, etc. The `level` and `label` orders must match.

```{r}
studentsB <- students %>% 
  mutate(Year = factor(Year, 
                       levels = c("Freshman",
                                  "Sophomore",
                                  "Junior",
                                  "Senior",
                                  "Graduate Certificate Program"),
                       labels = c("FR", 
                                  "SO", 
                                  "JR", 
                                  "SR", 
                                  "GR")))
```


## The Years Table One More Time

```{r}
levels(studentsB$Year)
studentsB %>% 
  count(Year)
```


## Alternative: Recoding Factors

Sometimes we want to rename or combine some factor levels. The `fct_recode` function from the `forcats` package allows us to do this. The function `fct_collapse()` is another alternative. This time we will create a new variable `Year2`.

```{r}
studentsC <- studentsA %>%
  mutate(Year2 = fct_recode(Year,
    FR = "Freshman",
    SO = "Sophomore",
    JR = "Junior",
    SR = "Senior",
    GR = "Graduate Certificate Program"
  ))
glimpse(studentsC$Year2)   # glimpse just one variable using $
```


## Combining Levels with `fct_recode()`

Assigning more than one level to the same new level name will combine the levels. Here all the majors besides `Mathematical Sciences` become one group.

```{r}
library(tidyverse) # or library(forcats)
studentsD <- studentsA %>%
  mutate(CodedMajor2 = fct_recode(CodedMajor,
    "Math" = "Mathematical Sciences",
    "Not Math" = "Computer Science",
    "Not Math" = "Other Sciences",
    "Not Math" = "Other"
  ))
glimpse(studentsD$CodedMajor2)   # glimpse just one variable using $
```


## Recoded Variable Table

```{r}
studentsD %>% 
  count(CodedMajor2) %>% 
  mutate(proportion = prop.table(n))
```

Notice that the levels are ordered as they were originally. Since `Computer Science` was the first category before, now `Not Math` is first because it contains `Computer Science`.

How could we fix this so that `Math` is the first row in the table?


## Use a Single `mutate()` Statement

You can combine different functions in one `mutate()` statement.

```{r}
studentsE <- students %>% 
  mutate(Year = factor(Year, 
                       levels = c("Freshman",
                                   "Sophomore",
                                   "Junior",
                                   "Senior",
                                   "Graduate Certificate Program")),
         Year2 = fct_recode(Year,
                            FR = "Freshman",
                            SO = "Sophomore",
                            JR = "Junior",
                            SR = "Senior",
                            GR = "Graduate Certificate Program")
  )
```


## Documentation Links

* [forcats](https://forcats.tidyverse.org/)

* [`fct_recode()`](https://forcats.tidyverse.org/reference/index.html)

* [`fct_collapse()`](https://forcats.tidyverse.org/reference/fct_collapse.html)

* [`factor()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/factor), or see a [quick DataCamp tutorial](https://www.datacamp.com/tutorial/factors-in-r)

* [`count()`](https://dplyr.tidyverse.org/reference/count.html)

* [`distinct()`](https://dplyr.tidyverse.org/reference/distinct.html)

* [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html)

* `prop.table()` is a base `R` function we are using with `count()` to calculate proportions from the column of counts

