---
title: "Measures of Variability"
subtitle: "Companion `R` Code for JB Statistics Video Lesson"
author: "Created by Jill E. Thomley, Appalachian State University"
date: "Last updated on `r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: TRUE
    toc_depth: 4
---

&#9654; [GO TO VIDEO](https://youtu.be/Cx2tGUze60s)

***

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

## Packages Used

```{r loadpackages}
library(tidyverse)
library(kableExtra)
```

## Datasets Used

### Sample of 4 Observations

```{r}
sample4 <- tibble(data = c(45, 64, 68, 51))
glimpse(sample4)
head(sample4)
```

### Guinea Pig Survival Data

```{r}
gp <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/JBStatistics/jbstatsguineapigs.csv")
glimpse(gp)
head(gp)
```

## Example with 4 Observations

### Steps in Calculating MAD and SD

Doing the calculations for deviations the long way. Deviations are the distances of each point from the mean. We use `mutate()` to create new columns from the original `data` column.

#### Calculate all the deviations

```{r}
devs <- sample4 %>%
  mutate(
    deviations = data - mean(data),
    absolute_devs = abs(deviations),
    square_devs = deviations^2
  )
devs
```

#### Sum the columns of deviations

```{r}
sums <- devs %>%
  summarize(
    n = n(),
    sum_devs = sum(deviations),
    sum_abs_devs = sum(absolute_devs),
    sum_sqr_devs = sum(square_devs)
  )
sums
```

#### Compute MAD and SD from sums

```{r}
stats <- sums %>%
  summarize(
    MAD = sum_abs_devs / n,
    s = sqrt(sum_sqr_devs / (n - 1))
  )
stats
```

### `R` Functions for Summarizing Data

We can do many of the same calculations directly using `R`'s built-in functions.

```{r}
sample4 %>%
  summarize(
    n = n(),
    xbar = mean(data),
    MAD = sum(abs(data - xbar)) / n,
    s = sd(data)
  )
```

## Example with Guinea Pig Survival Data

Summarize the overall dataset. Base `R` does not have a function for mean absolute deviation.

```{r}
gp_stats_all <- gp %>%
  summarize(
    n = n(),
    xbar = mean(survivaldays),
    s = sd(survivaldays),
    rng = diff(range(survivaldays))
  )
gp_stats_all
```

We can format the output for presentation using `kable()`.

```{r}
gp_stats_all %>%
  kable(digits = 2) %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 14
  )
```

Summarize the data by groups.

```{r}
gp_stats_grp <- gp %>%
  group_by(group) %>%
  summarize(
    n = n(),
    xbar = mean(survivaldays),
    s = sd(survivaldays),
    rng = diff(range(survivaldays))
  )

gp_stats_grp %>%
  kable(digits = 2) %>%
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 14
  )
```

## References

Tidyverse: [Create, modify, and delete columns](https://dplyr.tidyverse.org/reference/mutate.html)

Tidyverse: [Summarise each group to fewer rows](https://dplyr.tidyverse.org/reference/summarise.html)

[KableExtra Tutorial](https://rstudio-pubs-static.s3.amazonaws.com/444395_76727eaf9c774fa8bf932bed16500a00.html)

***
```{r}
sessionInfo()
```
