---
title: "Measures of Central Tendency"
subtitle: "Companion `R` Code for JB Statistics Video Lesson"
author: "Created by Jill E. Thomley, Appalachian State University"
date: "Last updated on `r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: TRUE
    toc_depth: 4
---

&#9654; [GO TO VIDEO](https://youtu.be/NM_iOLUwZFA)

***

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

## Packages Used

```{r loadpackages}
library(tidyverse)
library(kableExtra)
```

## Datasets Used

### Sample of 4 Observations

```{r}
# stored as a vector versus a data frame

sample6 <- c(1.4, 3.9, 5.0, 6.2, 8.4, 14.1)
glimpse(sample6)
```

### Guinea Pig Survival Data

```{r}
gp <- read_csv("https://raw.githubusercontent.com/STAT-JET-ASU/Datasets/master/JBStatistics/jbstatsguineapigs.csv")
glimpse(gp)
head(gp)
```

## Example with 6 Observations

### Original Observations

Since the data is a vector rather than a dataset, we are not using `summarize()`.

```{r}
(xbar <- mean(sample6))
(xmed <- median(sample6))
```

<details>
```{r}
# https://ggplot2.tidyverse.org/reference/geom_dotplot.html
# https://ggplot2.tidyverse.org/reference/scale_continuous.html
# https://ggplot2.tidyverse.org/reference/labs.html
# https://ggplot2.tidyverse.org/reference/geom_text.html
# https://ggplot2.tidyverse.org/reference/theme.html
# https://personal.sron.nl/~pault/#sec:qualitative

dot1 <- ggplot(NULL, aes(x = sample6)) +
  geom_hline(yintercept = 0) +
  geom_dotplot(
    binwidth = 1,
    dotsize = 0.5,
    color = "black",
    fill = "#DDAA33"
  ) +
  geom_text(aes(x = sample6, y = 0.1),
    label = sample6
  ) +
  geom_segment(aes(x = xbar, xend = xbar, y = -0.2, yend = 0),
    arrow = arrow(length = unit(0.3, "cm")),
    color = "#BB5566"
  ) +
  geom_text(aes(x = xbar, y = -0.25),
    label = paste("mean = ", xbar),
    fontface = "italic",
    color = "#BB5566",
    hjust = 0
  ) +
  geom_segment(aes(x = xmed, xend = xmed, y = -0.2, yend = 0),
    arrow = arrow(length = unit(0.3, "cm")),
    color = "#004488"
  ) +
  geom_text(aes(x = xmed, y = -0.25),
    label = paste("median = ", xmed),
    fontface = "italic",
    color = "#004488",
    hjust = 1
  ) +
  scale_x_continuous(
    limits = c(0, 27),
    breaks = seq(from = 0, to = 27, by = 1)
  ) +
  scale_y_continuous(breaks = NULL) +
  labs(
    title = "dotplot for example of 6 data points",
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```
</details>

```{r}
dot1
```

### Larger Maximum Value

Notice how the mean increases by 2 because of the new larger maximum, but the median does not.

```{r}
sample6mod <- c(sample6[1:5], 26.1)
glimpse(sample6mod)
```

```{r}
(xbar_mod <- mean(sample6mod))
(xmed_mod <- median(sample6mod))
```

## Example with Guinea Pig Survival Data

```{r}
gp %>%
  filter(group == "treatment") %>%
  ggplot(aes(x = survivaldays)) +
  geom_histogram(
    breaks = seq(from = 0, to = 600, by = 50),
    color = "darkgray",
    fill = "lightblue"
  ) +
  labs(
    title = "Survival times of 60 guinea pigs after being infected with tuberculosis",
    x = "Lifetime (days)",
    y = "Frequency"
  ) +
  theme_bw()
```

```{r}
ggplot(gp, aes(x = survivaldays)) +
  geom_histogram(
    breaks = seq(from = 0, to = 800, by = 50),
    color = "darkgray",
    fill = "lightblue"
  ) +
  facet_grid(group ~ .) +
  labs(
    title = "Survival Times of Guinea Pigs",
    subtitle = "control group vs. treatment group infected with tuberculosis",
    x = "Lifetime (days)",
    y = "Frequency"
  ) +
  theme_bw()
```

```{r}
(gp_stats_grp <- gp %>%
  group_by(group) %>%
  summarize(
    xbar = mean(survivaldays),
    xmed = median(survivaldays)
  ))

gp_stats_grp %>%
  kable(digits = 1) %>% # 1 decimal place
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 14
  )
```

## References

Tidyverse: [Create, modify, and delete columns](https://dplyr.tidyverse.org/reference/mutate.html)

Tidyverse: [Summarise each group to fewer rows](https://dplyr.tidyverse.org/reference/summarise.html)

Tidyverse: [Histograms and Frequency Polygons](https://ggplot2.tidyverse.org/reference/geom_histogram.html)

[KableExtra Tutorial](https://rstudio-pubs-static.s3.amazonaws.com/444395_76727eaf9c774fa8bf932bed16500a00.html)

***
```{r}
sessionInfo()
```
