---
title: "Understanding Z-Scores"
subtitle: "Companion `R` Code for Stats 101 Video Lesson"
author: "Created by Jill E. Thomley, Appalachian State University"
date: "Last updated on `r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: TRUE
    toc_depth: 4
---

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

&#9654; [GO TO VIDEO](https://youtu.be/bUu5HIHIrRw)

***

## Packages Used

```{r loadpackages}
library(tidyverse)
```

## Datasets Used

### Class 1 Scores

```{r}
class1 <- tibble(Score = sort(c(85, 95, 75, 80, 90)))
```

### Class 2 Scores

```{r}
class2 <- tibble(Score = sort(c(88, 79, 91, 85, 82)))
```

## Calculations

### Summary Statistics

Class 2 has the same mean but a smaller standard deviation than Class 1.

```{r}
class1 %>%
  summarize(
    mean_class1 = mean(Score),
    sd_class1 = sd(Score)
  )
class2 %>%
  summarize(
    mean_class2 = mean(Score),
    sd_class2 = sd(Score)
  )
```

### Z-Scores for Class 1

This example works out the math like the table in the video. `Difference` is the non-standardized distance from the mean.

```{r}
class1 %>%
  mutate(
    Mean = mean(Score),
    Difference = Score - Mean,
    Z = Difference / sd(Score)
  )
```

We can do this directly using the `scale()` function in `R`. If you look at `class1_zscores` in the Environment pane, you will see the mean and standard deviation saved as *attributes* of the Z column (properties explain what an object represents and how it should be interpreted by R --- in this case the values used for scaling). 

```{r}
class1_zscores <- class1 %>%
  mutate(Z = scale(Score))
class1_zscores
attributes(class1_zscores$Z)
```

### Z-Scores for Class 2

Notice that the z-scores here are the same as Class 1, but the individual scores are closer to the mean than Class 1 (`Difference` is smaller). This is because the standard deviation (our unit of measure or "ruler") is smaller in Class 2.

```{r}
class2 %>%
  mutate(
    Mean = mean(Score),
    Difference = Score - Mean,
    Z = Difference / sd(Score)
  )
```

```{r}
class2_zscores <- class2 %>%
  mutate(Z = scale(Score))
class2_zscores
```

```{r}
cbind(class1_zscores, class2_zscores)
```

***
```{r}
sessionInfo()
```
